---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(SomaEnrich)
library(dplyr)

ver <- desc::desc_get_version(".")
ver <- paste0("https://img.shields.io/badge/Version-", ver,
              "-success.svg?style=flat&logo=github")
```


# SomaEnrich


<!-- badges: start -->
<!-- badges: end -->

_This R package is an in-development resource and is research use only._

## Overview

The `SomaEnrich` R package provides tools to perform pathway analysis with 
proteomic data obtained from the SomaScan assay. The package also provides 
auxiliary functions for manipulating, converting, and filtering data prior
to pathway analysis, as well as detailed documentation describing the unique 
steps required to perform these operations with proteomic data (as opposed 
to genomic data).

For more information about SomaScan data files (i.e. the ADAT, or `.adat` file) 
and their format, please see the 
[SomaLogic-Data](https://github.com/SomaLogic/SomaLogic-Data) GitHub repository.

If you run into any issues/problems with `SomaEnrich` or have questions 
about SomaScan analysis that are not answered by this package, please
consult the [issues](https://github.com/SomaLogic/SomaEnrich/issues/)
page and submit an issue and/or feature request.

### Package Key Features

- **Overrepresentation Analysis (ORA)**: Identify pathways enriched in a set of 
  differentially expressed proteins using hypergeometric testing
- **Gene Set Enrichment Analysis (GSEA)**: Perform preranked GSEA to detect 
  coordinated changes in predefined gene sets
- **SomaScan Identifier Handling**: Conversion between common gene identifiers 
  and SomaScan-specific `AptNames`
- **Visualization Tools**: Plots to visualize the results of ORA and GSEA

---------------------

## Installation

_Note: SomaEnrich was developed and tested on macOS and Linux operating_
_systems. Users running Windows may encounter unexpected behavior or errors._

`SomaEnrich` requires that the user have R >= 4.4.0 installed on their
system. 

To install and load `SomaEnrich` from this repository, please run the 
following:

```{r install-github, eval=FALSE}
remotes::install_github("SomaLogic/SomaEnrich")
```

Once installed, the package can be loaded using the usual syntax:

```{r load-pkg, eval=FALSE}
library(SomaEnrich)
```


## Package Dependencies

Below are the package dependencies for `SomaEnrich` (see also the DESCRIPTION file):

* [SomaDataIO](https://CRAN.R-project.org/package=SomaDataIO) - install from CRAN: `install.packages("SomaDataIO")`
* [fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html) - install from Bioconductor: `BiocManager::install("fgsea")`
* [ggplot2](https://CRAN.R-project.org/package=ggplot2) - install from CRAN: `install.packages("ggplot2")`
* [patchwork](https://CRAN.R-project.org/package=patchwork) - install from CRAN: `install.packages("patchwork")`
* [rlang](https://CRAN.R-project.org/package=rlang) - install from CRAN: `install.packages("rlang")`
* [scales](https://CRAN.R-project.org/package=scales) - install from CRAN: `install.packages("scales")`
* [stringr](https://CRAN.R-project.org/package=stringr) - install from CRAN: `install.packages("stringr")`
* [tidyr](https://CRAN.R-project.org/package=tidyr) - install from CRAN: `install.packages("tidyr")`
* [withr](https://CRAN.R-project.org/package=withr) - install from CRAN: `install.packages("withr")`

The GSEA and ORA functions in this package use the respective algorithms from 
`fgsea`, under the hood. Please see the `fgsea` documentation for more 
information.

---------------------

## Quick Start

### Identifier Conversion

Convert SomaScan analytes to genes:

```{r convert-to-genes}
head(ex_apt_pathway)

apt2gene(head(ex_apt_pathway))
```

Convert genes to SomaScan analytes:

```{r convert-to-apts}
genes <- withr::with_seed(123, sample(ex_gene_pathway, 5))

gene2apt(genes) # Returns NA if no analyte is found
```

Convert a GO term from genes to SomaScan analytes:

```{r convert-go-term}
go2apt("GO:0022602")
```


### Overrepresentation Analysis (ORA)

Identify pathways represented in a set of differentially expressed proteins:

```{r ora-example}
# Get top differentially expressed genes from a pre-sorted results table
deg <- head(t_tests$EntrezGeneSymbol, 50)
uni <- t_tests$EntrezGeneSymbol

# Run ORA using GO Biological Process (the default)
ora_res <- somaORA(features = deg, universe = uni)
head(ora_res)
```

```{r ora-viz, fig.width=8, fig.height=7}
# Visualize the ORA results
plotBubble(ora_res, n_pathways = 15)
```


### Gene Set Enrichment Analysis (GSEA)

Pre-ranked GSEA can be performed using the following workflow:

```
prepareRanks() --> somaPrGSEA() --> plotES()
```

#### Preparing Ranked Input

Resolve many-to-one analyte-gene mappings and split heterodimers to create
a clean, unique ranking vector suitable for GSEA:

```{r prepare-ranks}
unique_ranks <- prepareRanks(stats = t_tests$log2_fc,
                             features = t_tests$EntrezGeneSymbol,
                             resolve_multimapping = TRUE,
                             resolve_method = "abs")

head(unique_ranks)
```

#### Running GSEA

Perform preranked GSEA to detect coordinated pathway changes:

```{r gsea-example}
# Use GO Biological Process (the default)
gsea_res <- somaPrGSEA(ranks = unique_ranks)
```

```{r gsea-viz, fig.width=8, fig.height=6}
# Visualize the enrichment score for a chosen pathway
plotES("female gamete generation", gsea_results = gsea_res)

# Add leading edge points to plot
plotES("female gamete generation", gsea_results = gsea_res, show_leading_edge = TRUE)
```


---------------------

## Package Data Objects

`SomaEnrich` comes bundled with data objects for use in examples and vignettes:

| Object | Description |
|--------|-------------|
| `example_data_11k` | A `soma_adat` containing example 11K SomaScan data |
| `t_tests` | Results of a two-group comparison of the Sex variable to identify genes differentially expressed in males vs. females |
| `pathway_map` | Mapping of genes to pathways from GO and MSigDB |
| `ex_gene_pathway` | Example pathway comprised of gene identifiers |
| `ex_apt_pathway` | Example pathway comprised of SomaScan `AptNames` |

The GO and MSigDB data in `pathway_map` is made available under the terms of 
the [CC BY 4.0 license](https://creativecommons.org/licenses/by/4.0/). See 
`?pathway_map` for more details. 


---------------------

## MIT LICENSE

`SomaEnrich` is licensed under the MIT license and is intended
solely for research use only (RUO) purposes. The code contained
herein may not be used for diagnostic, clinical, therapeutic, or other
commercial purposes.

- See:
  - [LICENSE](https://github.com/SomaLogic/SomaEnrich/blob/main/LICENSE.md)
- The MIT license:
  - <https://choosealicense.com/licenses/mit/>
  - [https://www.tldrlegal.com/license/mit-license/](https://www.tldrlegal.com/license/mit-license)
- Further:
  - "SomaEnrich" and "SomaLogic" are trademarks owned by Illumina, Inc. No license is hereby granted to these trademarks other than for purposes of identifying the origin or source of this Software.
