---
title: "Non-1:1 Relationships in the SomaScan Menu"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Non-1:1 Relationships in the SomaScan Menu}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SomaDataIO)
library(SomaEnrich)
library(dplyr)
library(tibble)

ex_gene <- "KLK3"
ex_gene_name <- "Prostate-Specific Antigen"
rank_metric <- "t_stat"

t_tests <- dplyr::arrange(t_tests, desc(abs(t_stat)), fdr)

gene_ranked <- filter(t_tests, EntrezGeneSymbol == ex_gene) |>
    select(AptName, EntrezGeneSymbol, !!sym(rank_metric)) |>
    arrange(desc(!!sym(rank_metric)))
```

_This R package is an in-development resource and is research use only._

-----

# Overview

The SomaScan menu contains instances of non-1:1 relationships between
SOMAmer<sup>&reg;</sup> Reagents and their target proteins. In simple terms, 
this means there is sometimes not a perfect, singular line that can be drawn 
between one SOMAmer Reagent, its target protein, and the gene that transcribes 
it. This is a unique characteristic of SomaScan, stemming from the biological 
complexity and diversity of the proteome. Understanding and 
handling these non-1:1 relationships can be a crucial 
processing step for pathway analysis, particularly when using 
tools or methods developed for gene-based (transcriptomic) data. 

This vignette will explain why these relationships are relevant to pathway 
analysis, and demonstrate how the tools in SomaEnrich can be used to streamline
a pathway analysis workflow.


---------------

## How SomaScan Measures Proteins

To better understand non-1:1 relationships in the SomaScan menu, and how their 
presence shapes data analysis, it first helps to know how the assay works. 

SOMAmer Reagents bind to three-dimensional protein structures, like epitopes, 
and are conceptually similar to monoclonal antibodies in immunoassays. Because 
SomaScan measures specific structural features, rather than entire proteins, a 
single protein can be detected by multiple SOMAmer Reagents if it contains 
multiple distinct binding structures. 

So, for example, the following 1:many pattern could be observed in the data:

    Protein 1 ____ SOMAmer Reagent 1
            |_____ SOMAmer Reagent 2
            |_____ SOMAmer Reagent 3

This differs from non-affinity-based protein measurement technologies like 
mass spectrometry, where measurements typically correspond more directly to 
individual proteins or genes.

With this context in mind, the remainder of this vignette will use example 
SomaScan data and R code to show what non-1:1 relationships look like in 
the data, how they impact gene-based analyses like GSEA, and options
for handling them.

---------------

## Load R Packages

```{r load-pkgs}
library(SomaDataIO)
library(SomaEnrich)
library(dplyr)
library(tibble)
```

## Prepare for GSEA

In pre-ranked GSEA, an ordered statistical metric is required, often
derived from the results of differential expression analysis. This can be 
pulled from the `SomaEnrich::t_tests` data set. The `r rank_metric` will be 
used as the ranking metric - please note that this may not be the best metric 
for your experiment, and is merely used here as an example.

Entrez gene symbols will be used as vector names, to match the gene set
networks already available in SomaEnrich. Below is a subset of the `t_tests`
data set, sorted by |`r rank_metric`|:

```{r t-tests-example}
t_tests |>
    select(SeqId, EntrezGeneSymbol, t_stat) |>
    arrange(desc(abs(t_stat))) |>
    head(n = 10)
```

Every `SeqId` in the above table is unique. However, some genes, like
`r ex_gene_name` (`r ex_gene`), appear multiple times in the
"EntrezGeneSymbol" column. This illustrates a 1:many (i.e. non-1:1)
relationship between `r ex_gene` and `SeqIds` in the data.

If Entrez gene symbols are solely used to identify each data point in the
ranked vector, an input vector for GSEA could look like this:

```{r ranks-duplicates}
rank_vec <- t_tests$t_stat
names(rank_vec) <- t_tests$EntrezGeneSymbol

rank_vec[1:10]
```

In the top example entries alone, some gene names are
seen more than once, and combined gene entities with a "|" character are
present in the vector names. Both of these patterns are unique to
SomaScan-derived data and can complicate the GSEA workflow. 

In particular, the presence of duplicate entries may have unintended
consequences for GSEA, if not corrected, as it violates one of the core
assumptions of GSEA.


## Assumptions of GSEA

A critical assumption of the GSEA algorithm is that 1:1 mapping is
present between ranked entries and biological entities, and that each gene in
the ranked list is only represented once [@gsea-paper; @gsea-manual].
Violations of this assumption can disproportionately weight genes during
calculation of the enrichment score, resulting in the following undesirable
effects:

* Hits (highly-ranked genes found in a gene set) could be counted multiple times
* Incidences of multiple counting could create inflated enrichment scores
* Enrichment scores for small gene sets could be especially distorted

The non-1:1 gene:SOMAmer Reagent relationships in the SomaScan menu
stress this assumption when using SomaScan-derived ranking statistics for
GSEA.


## SOMAmer Reagent Targets

In a SomaScan data set, SOMAmer Reagents and `SeqIds` have a
1:1 relationship, where each `SeqId` uniquely identifies a single SOMAmer
reagent. However, as described earlier in this document, *proteins* do not 
necessarily have a 1:1 relationship with a
single SOMAmer Reagent. In these cases, external identifiers like the UniProt 
ID (or gene ID) may not fully distinguish the exact target structure of a 
SOMAmer Reagent.

For example, as observed earlier, the `r ex_gene` gene appears in the 11K
SomaScan menu multiple times:

```{r ex-gene-menu, echo=FALSE}
ex_gene <- "KLK3"

SomaDataIO::getAnalyteInfo(example_data_11k) |>
    filter(EntrezGeneSymbol == ex_gene) |>
    select(SeqId:EntrezGeneSymbol, -SeqIdVersion, -SomaId, -EntrezGeneID)
```

All 3 `SeqIds` in the above table are unique, but the other values are (for
the most part) identical. In the `example_data_11k` ADAT, all these
unique analytes are present, each with a distinct RFU measurement:

```{r ex-gene-rfu, echo=FALSE}
ex_analytes <- SomaDataIO::getAnalyteInfo(example_data_11k) |>
    filter(EntrezGeneSymbol == ex_gene) |>
    pull(AptName)

select(example_data_11k, SampleId, all_of(ex_analytes)) |>
    SomaDataIO::rm_rn() |>
    head(n = 10) |>
    tibble::tibble()
```

In `t_tests`, multiple analytes targeting `r ex_gene` were (after sorting)
located near the bottom of the ranked gene vector.

So, let's return to the example vector we made for pre-ranked GSEA. Proceeding
with this vector as-is may have the following unintended effects:

* `r ex_gene` appears multiple times in the ranked vector, potentially leading to over-representation of `r ex_gene` in the final GSEA results
    * This effect applies to any other duplicated entities in the vector
* The vector elements with "|"-delimited names will not match gene symbols in the available GO or MSigDB resources, and consequently will be counted as a "miss" in the enrichment score calculation, rather than a hit
    * This can result in inaccurate counting of these highly-ranked entities in the final enrichment results

Solutions for both of these points are implemented by SomaEnrich, and will be
described in greater detail below.

---------------


# Solutions in SomaEnrich

The functions in SomaEnrich automatically handle instances of 1:many 
mapping between SOMAmer Reagents (and therefore, SomaScan measurements) 
and external gene or protein identifiers.

SomaEnrich provides functions that perform:

1. Data cleaning operations, to split apart delimited identifiers that signify protein heterodimers
1. Measurement filtering and selection, to resolve non-1:1 mapping for gene-based analyses like GSEA

The following sections will explain how functions in SomaEnrich handle these
scenarios.


## Working with Heterodimers and "|" Delimiters

The SomaScan menu contains instances where a "|" character is
used to delineate two (or more) gene or protein identifiers. This notation
indicates that a given SOMAmer Reagent binds a heterodimer protein or larger
protein heterocomplex, where the protein subunits originate from different 
genes. To fully capture this information in the menu, all genes with protein 
products in the complex are separated by a
delimiter character, typically "|".

However, the biological knowledge bases typically used by popular pathway 
analysis tools, like GO and KEGG, were developed for gene-centric data, and do 
not contain this type of annotation. 

When performing GSEA, all identifiers used for analysis inputs should be of
consistent type, and in the same format. By default, SomaEnrich 
uses Entrez gene symbols for both GO and MSigDB gene sets; therefore, the gene
symbols in the ranked input vector must be compatible Entrez gene symbols.
Because this heterodimer gene notation is unique to SomaScan, genes delimited
by "|" will not be compatible with the entries in GO and MSigDB. Consequently,
any gene in the input vector not found in the gene set will
contribute to the negative increment of the running sum statistic, but will
not be removed from the data set [@gsea-paper]. This can affect
the calculation of the enrichment score - if a highly-ranked gene _should_
contribute to the running sum, but can't be correctly matched to any member of
the gene set, the enrichment contribution from that gene will not be adequately
captured.

To ensure consistency in gene symbols across both input ranks and gene sets,
delimited identifiers are split prior to analysis (see the `split_ids` argument
of `SomaEnrich::somaPrGSEA()`).

Here's a preview of the first 10 rows of data, without the splitting operation:

```{r gene-symbols}
# Prior to split
t_tests |>
    select(AptName, EntrezGeneSymbol, t_stat) |>
    head(n = 10)
```

And here is that same data post-split:

```{r split-gene-symbols}
t_tests_split <- t_tests |>
    select(AptName, EntrezGeneSymbol, t_stat) |>
    mutate(row_id = row_number()) |>
    tidyr::separate_rows(EntrezGeneSymbol, sep = "\\|")

head(t_tests_split, n = 12)
```

This splitting procedure can have the following effects:

1. Splitting the heterodimers requires duplicating the ranking metric value, which can introduce ties to the ranked vector
1. Splitting the heterodimers can contribute to instances of gene name duplication in the ranked input vector

Let's use the table above to better illustrate the first point. The following
analyte, `seq.4914.10`, maps to 3 Entrez gene symbols:

```{r ex-heterodimer}
t_tests |>
    dplyr::select(AptName, SeqId, Target, EntrezGeneSymbol, t_stat) |>
    filter(AptName == "seq.4914.10")
```

The string `CGA|CGB3|CGB7` is unlikely to be matched in any gene-based database,
so the string must be split into its constituent parts:

```{r splitting-heterodimers}
t_tests |>
    dplyr::select(AptName, SeqId, Target, EntrezGeneSymbol, t_stat) |>
    filter(AptName == "seq.4914.10") |>
    tidyr::separate_rows(EntrezGeneSymbol, sep = "\\|")
```

Now, if we were to use the `r rank_metric` as input into `somaPrGSEA()`,
the input would look like this:

```{r inputs-with-ties, echo=FALSE}
t_tests |>
    dplyr::select(AptName, EntrezGeneSymbol, t_stat) |>
    filter(AptName == "seq.4914.10") |>
    tidyr::separate_rows(EntrezGeneSymbol, sep = "\\|") |>
    dplyr::select(-AptName) |>
    tibble::deframe()
```

The genes are unique, but the actual values used for ranking the genes are the
same. This can produce a warning during GSEA, stating that the ranking vector
has ties. Essentially, these 3 elements can't be ranked properly, because they
all have the same ranking value. While this is typically undesirable if it
occurs in the raw, unmodified ranking vector, the decision to split
heterodimers during pre-processing can itself create ties. This is _expected_
behavior when splitting SomaScan heterodimers.


## Multiple Targets Map to One Gene

Let's move on to the second unintended effect of splitting heterodimers: it can
contribute to instances of gene name duplication in the ranked input vector.
This may result in >1 value per `AptName`, with a unique gene symbol
for each value. Take `3032-11`, for example; prior to the split operation, it
possesses one `r rank_metric` value:

```{r single-aptname}
filter(t_tests, AptName == "seq.3032.11") |>
    select(AptName, EntrezGeneSymbol, t_stat)
```

Post-split, the `r rank_metric` value for `3032-11` is split across multiple
rows.

```{r split-aptname}
filter(t_tests_split, AptName == "seq.3032.11")
```

Because genes like CGA can be mapped to >1 SOMAmer Reagent, the
above step produces multiple `r rank_metric` values for CGA, when viewed in
combination with the rest of the data set:

```{r single-gene}
filter(t_tests_split, EntrezGeneSymbol == "CGA")
```

All of these values are associated with the gene CGA, but it's best practice
to not use this vector as-is for GSEA. To avoid duplication of gene 
identifiers, only one `r rank_metric` value for CGA will be chosen. But why
only choose one, and how should it be selected?


## Selecting One Measurement per Gene

In transcriptomics, it's common practice to combine transcript or probe-based
measurements and sum them to obtain gene-level counts.
When performing GSEA with SomaScan data, it may be tempting to apply this
methodology and, for example, sum together all measurements from analytes
mapping to CGA or `r ex_gene`. This would produce one single measurement for
each gene and resolve the identifier duplication problem described previously.

However, it is _not_ advised to collapse or aggregate analytes into a single
entity based on their protein target or gene. Instead, it is standard practice
to keep SomaScan measurements **distinct** whenever possible.

If each SOMAmer Reagent is measuring a different isoform or region of a
protein, collapsing these data points may lose valuable information about those
protein regions. Instead, it is recommended to treat each analyte as an
independent measurement, and select a single "best" measurement to
be _representative_ of each protein (or, in this case, gene) when multiple
measurements are present.

But what constitutes the "best" measurement? This can can vary
depending on experiment and ranking metric used for pre-ranked GSEA.
SomaEnrich provides a few different methods to make this selection.

Let's return to the `r ex_gene` example; we will sort the `r ex_gene` analytes
by our chosen ranking metric (`r rank_metric`):

```{r show-gene-measurements}
filter(t_tests, EntrezGeneSymbol == ex_gene) |>
    select(AptName, EntrezGeneSymbol, t_stat) |>
    arrange(desc(t_stat))
```

The "top" `r ex_gene`-associated analyte (sorted by `r rank_metric`) is
`r gene_ranked$AptName[1]`. If we follow the recommendation to retain one
measurement per gene and arbitrarily choose the first occurrence, only this
`r rank_metric` value for `r ex_gene` will be used for input into GSEA, and the
others will be lost. However, this can have unintended consequences. What if
we're interested in the value with the largest _magnitude_, regardless of sign?
Or what if the selected ranking metric is a p-value? The same sorting method
may not work for both p-values (where the smallest value is preferred) and fold
change values (where the largest absolute value is preferred).

Merely sorting in ascending/descending order and choosing the first occurrence
may not return the expected result.


## Selection Methods in SomaEnrich

SomaEnrich performs this sorting and selection step under the hood in
`somaPrGSEA()` via the `resolve_multimapping` argument. When
`resolve_multimapping = TRUE` (the default), the function will create a
catalog of all unique genes present in a ranked data frame, identify all
SOMAmer Reagents associated with each gene, and select a single analyte based
on a user-defined criteria. By default, the statistic with the largest absolute
value is retained. This ensures that each protein target/gene is only
represented once in the data set, **without** collapsing values to create a
single summed measurement. 

However, like mentioned previously, this sorting method may not be
appropriate for all ranking metrics. The following measurement sorting and
filtering methods are supported:

- Filtering by **absolute value**  (the default): the largest absolute value is chosen (ignores sign)
- Filtering by **maximum**: the maximum value is chosen
- Filtering by **minimum**: the minimum value is chosen
- Filtering by **rank**: the smallest rank is chosen using pre-sorted data

For more complex sorting (i.e. using two or more sorting criteria), the "rank"
method may be more appropriate. After sorting by the desired criteria, like a
combination of FC and p-value in a single-sided test, a rank is assigned to the
input values, using the order in which they were sorted in. The first value is
ranked 1. In cases of multimapping, the value with the _lowest_ rank is
retained.

Please note that selection by "rank" and "minimum" can produce _different_
results! Filtering by rank selects based on position in the input vector
(given pre-sorting), while filtering by minimum selects based on actual numeric
value, and can be impacted by the presence of negative values when the metric
is signed. Similarly, depending on how the pre-sorting is done, "maximum" and
"rank" can select either the same or different values. Ultimately, filtering
by rank may lose information - the magnitude of difference in values is
ignored, and proximity to the start of the vector is prioritized.
Alternatively, if selecting by absolute value, magnitude of signal is
prioritized, while directionality (i.e. sign) is lost.

The chunk below shows how the available sorting options can identify
different values for fictional `GENE_A`:

```{r}
gene_a <- c(GENE_A = 0.01, GENE_A = -0.5, GENE_A = 5, GENE_A = -10)

max(gene_a)

min(gene_a)

max(abs(gene_a))

sort(gene_a, decreasing = TRUE)[[1]] # "rank" option selects the 1st value when sorted
```

The code above highlights the importance of assessing which selection
method is most appropriate for your data. Multiplied over hundreds of cases of
non-1:1 mapping, the chosen measurement filtering method can have a notable
impact on GSEA data.

After applying the filtering technique described above to the
example rank vector, the results should look like this:

```{r clean-vector, echo=FALSE}
final_rank_vec <- prepareRanks(stats = t_tests[[rank_metric]],
                               features = t_tests$EntrezGeneSymbol,
                               verbose = FALSE)

print(head(final_rank_vec, n = 10))
```

No duplicate genes or "|"-delimited heterodimers remain in the preview of the
ranking vector.

This filtering procedure is performed for all genes in the input ranked
vector, and the filtered vector will then be used as input for GSEA.

The output of `somaPrGSEA()` will include this filtered and modified vector,
for reference. 


-------------

# Summary

* The SomaScan menu contains **non-1:1 relationships** between SOMAmer Reagents and their target proteins/genes, which can complicate gene-centric analyses like GSEA
* **Delimited identifiers** (e.g., "GENE1|GENE2") must be split so each symbol can be properly matched to gene sets
* **SomaLogic recommends keeping analyte measurements distinct** rather than collapsing or aggregating them by gene — this preserves valuable information about protein isoforms and regions
* SomaEnrich automatically handles non-1:1 relationships by **selecting a single "best" measurement** per gene, rather than summing measurements together
* **"Best"** measurement can be defined in various ways, and a few options are provided to do this automatically
* The R functions `prepareRanks()`, `somaPrGSEA()`, and `somaORA()` are capable of performing these processing steps

Please keep in mind that these processing steps are an *optional* feature, and
ultimately you must decide if they are appropriate for your analysis.

-------------

# Questions

If you have any additional questions about the SomaScan menu, we are here to
help. Please reach out to us via:

* GitHub [issues](https://github.com/SomaLogic/SomaEnrich/issues)
* Global Scientific Engagement Team: <techsupport@somalogic.com>
* General SomaScan inquiries: <support@somalogic.com>


-------------

# References

