---
title: "Introduction to SomaEnrich"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Introduction to SomaEnrich}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(SomaEnrich)
library(SomaDataIO)
library(dplyr)
library(fgsea)
library(limma)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

t_tests <- dplyr::arrange(t_tests, desc(abs(t_stat)), fdr)

ex_gene <- "KLK3"
ex_gene_name <- "Kallikrein-related peptidase 3"
rank_metric <- "log2_fc"
```

_This R package is an in-development resource and is research use only._

-----

The `SomaEnrich` R package provides tools to perform standard pathway enrichment
analyses, like over-representation analysis (ORA) and gene set enrichment
analysis (GSEA), with SOMAmer<sup>&reg;</sup> Reagent-based SomaScan data. 

Why create this tool when there are so many pre-existing pathway analysis 
resources? The currently available ecosystem of software and R packages were 
often designed for (and derived from) transcriptomic data, and may not be as 
readily compatible with SomaScan data. SomaScan is a proteomics technology, and 
the SomaScan data file (`.adat`) is centered around SomaScan assay analytes 
(SOMAmer Reagents). This means that SomaScan measurements are representative of
SOMAmer Reagent *protein* targets, rather than transcribed genes, as they would
in transcriptomic data.

The tools in this package account for SomaScan-specific analysis complexities
that stem from this key difference in technologies.


---------------

# Main Features

The functions in this package can be divided into a few categories:

* Identifier mapping utilities
    + map SomaScan analytes to gene identifiers, and vice versa
    (`apt2gene()` & `gene2apt()`)
    + convert entire pathways of genes to SomaScan identifiers
    (`genePath2aptPath()` and `aptPath2genePath()`)

* Data wrangling resources
    + convert an `AptName`-centric `data.frame` into a gene-centric
    `data.frame` (`collapseAptData()`)
    + convert a data frame into a list of vectors, with one list element per
    unique group (`df2list()`)

* Enrichment analysis
    + prepare SomaScan statistical data for gene-based enrichment analysis (`prepareRanks()`)
    + perform pre-ranked gene set enrichment analysis (GSEA) using the `fgsea`
    algorithm (`somaPrGSEA()`)
    + perform over-representation analysis (ORA) using a hypergeometric test
    (`somaORA()`)

* Data visualization tools
    + create an enrichment plot from a GSEA results table (`plotES()`)
    + create a bubble plot from an ORA results table (`plotBubble()`)

Below is a basic walkthrough illustrating how SomaEnrich functions can be
used to perform pre-ranked GSEA, using differential expression results from
SomaScan assay data.


### A Note on ORA vs. GSEA

Both ORA and GSEA are widely used approaches for pathway enrichment, and both
are supported by SomaEnrich. In general:

- ORA tests whether a pre-selected set of features (e.g. differentially
expressed analytes above some significance or fold-change threshold) overlaps
with known pathways more than would be expected by chance. The process is
relatively straightforward: define a threshold, identify features that pass it,
and run a hypergeometric test. However, ORA requires an arbitrary cutoff to
define the feature list. Because of this requirement, ORA can be sensitive to
the choice of threshold.

- GSEA (in this case, pre-ranked) operates on a ranked list of _all_ measured
features, without applying a significance cutoff. Pre-ranked GSEA evaluates
whether the members of a given gene set tend to cluster toward the top or
bottom of a ranked list.

This vignette primarily focuses on GSEA because it requires
more considerations and pre-processing steps for SomaScan specifically. Please 
reference `?somaORA` for examples of performing ORA with SomaScan-derived
differential expression data.


---------------

# Example Workflow

## Getting Started

First, we need to load the R packages required for this vignette:

```{r load-pkgs, results='hide'}
library(dplyr)
library(fgsea)
library(limma)
library(SomaDataIO)
```

A requirement of pre-ranked GSEA is a numeric vector of statistical results
that can be used to rank genes. This workflow assumes that some ranking
procedure, like differential expression analysis, has already been performed
using data from a pre-processed SomaScan ADAT. Example ADATs can be obtained
from the 
[SomaLogic-Data](https://github.com/SomaLogic/SomaLogic-Data) GitHub
repository.

For more information about pre-processing your ADAT, please see the
[SomaDataIO Pre-Processing article](https://somalogic.github.io/SomaDataIO/articles/pre-processing.html).

For an example differential expression workflow (via t-test), please see the
[SomaDataIO Two-Group Comparison workflow](https://somalogic.github.io/SomaDataIO/articles/stat-two-group-comparison.html).

A differential expression results data set, called `t_tests`, is available as a
data object in `SomaEnrich`. The `t_tests` data set will be used as the
starting point for this analysis. `t_tests` was created by comparing two
groups, males and females, using the `Sex` variable from the
`SomaEnrich::example_data_11k` ADAT. See `?t_tests` for more information.

```{r preview-t-tests}
head(t_tests, n = 10)
```

Because `Sex` was used to compare these groups, a number of sex-associated
proteins are seen at the top of the t-test results, when sorted by `|t_stat|`.


## Selecting Ranking Metrics

SomaEnrich performs GSEA via the `somaPrGSEA()` function, which uses the
algorithm from the 
[fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html)
R package under the hood. Like fgsea,
`SomaEnrich::somaPrGSEA()` requires an input statistic to rank the features.

Any numeric statistic can be used as a ranking metric, including
test statistics, measures of significance (-log10(p-value)), or
signed/unsigned metrics measuring association of phenotype to analyte (fold
change or log2 fold change). However, because the statistic will be sorted in
descending order during calculation of the enrichment score, metrics in which
higher values = stronger signal are often used. The best ranking metric depends
on your data, experiment, and/or analysis question. See
[GSEA metrics for ranking genes](https://docs.gsea-msigdb.org/#GSEA/GSEA_User_Guide/#metrics-for-ranking-genes)
for more details.

For the purposes of this vignette, we will use the `r rank_metric` as a ranking
variable. Please note that this may not be the best metric for your experiment,
and is merely used here as an example.

We can create the GSEA input rank vector from the `t_tests` object,
using the `r rank_metric` column with values calculated for each analyte:

```{r create-rank-df}
rank_vec <- t_tests$log2_fc

names(rank_vec) <- t_tests$EntrezGeneSymbol

rank_vec[1:10]
```

Gene symbols/identifiers _or_ SomaScan `AptNames` may be used as names for the
ranking statistic vector. However, additional steps may be required when using
`AptNames`, and these will be described in a later section of this vignette.


## Resolving Non-1:1 Relationships

You may have noticed that the names of the `rank_vec` object above contain
duplicate values; for example, `r ex_gene` (`r ex_gene_name`) is represented
more than once in the first 10 elements of the vector, and CGA (glycoprotein
hormones alpha polypeptide) is represented multiple times in different 
"|"-delineated strings. If you are accustomed to working with gene-level data, 
this duplication may seem strange. These are instances of non-1:1 mapping 
(i.e. one identifier maps to >1 values) between
SOMAmer Reagents (with `SeqId` identifiers) and canonical protein or
gene target identifiers.

There are a variety of reasons why these entities may not have a 1:1
relationship. In some instances, a SOMAmer Reagent binds a heterodimer protein
(or larger heterocomplex), where protein subunits originate from different
genes. In other cases, multiple SOMAmer Reagents were developed for the same
protein target, or multiple proteoforms have a common gene product identifier.

Take `r ex_gene` as an example:

```{r ex-gene-targets}
filter(t_tests, EntrezGeneSymbol == "KLK3") |>
  dplyr::select(AptName:UniProt, log2_fc)
```

Multiple SOMAmer Reagents were developed for the `r ex_gene` protein target, as
seen in the table above for `r ex_gene`. All SOMAmer Reagents have a
unique SomaScan-specific identifier (`SeqId`), but in this case multiple
reagents possess the same gene and protein identifiers. This means that, if
we were only to use the "EntrezGeneSymbol" and "log2_fc" information to make a
named input vector to GSEA, we would see duplicate gene names in the vector.

The GSEA algorithm requires that the input ranking vector be comprised of only
**unique** values. This necessitates an extra pre-processing step to resolve
the duplication described here. 

Many functions in SomaEnrich select only **one** assay analyte to
represent each gene in enrichment analyses. The ideal method to select this
analyte can vary, depending on the ranking metric used; by default, the analyte
with the largest |`r rank_metric`| would be selected. This ensures that the
measurement from the analyte with the greatest association to the biological
feature of interest is retained.

This processing step is handled by the `prepareRanks()` function, which is also
called under the hood by `somaPrGSEA()`. This important step allows
proteomic SomaScan data to be adapted for gene-based analysis
workflows. For more information about this procedure, and for more detailed
descriptions of these non-1:1 relationships, please reference the other package
vignette via `browseVignettes("SomaEnrich")`.

With `prepareRanks()`, we can create a unique statistical vector that is
suitable for input to GSEA:

```{r prep-ranks, message=TRUE}
prepped_ranks <- prepareRanks(stats     = t_tests$log2_fc,
                              features  = t_tests$EntrezGeneSymbol,
                              split_ids = TRUE,            # Default
                              resolve_multimapping = TRUE, # Default
                              resolve_method = "abs")      # Default

head(prepped_ranks, n = 10)

# Only 1 KLK3 value remaining
which(names(prepped_ranks) == "KLK3")

# No gene name duplications
any(grepl("\\|", names(prepped_ranks)))
```

Only one value is present for gene KLK3, and all heterodimers or complexes have 
been split into their constituent parts. The `prepped_ranks` vector is ready for
pre-ranked GSEA.

As mentioned previously, this same function is called under the hood in
`somaPrGSEA()`, but `prepareRanks()` enables easy access to this processed
ranking metric _without_ performing GSEA, and the output can be used for other
gene-based analyses.


## Running Pre-Ranked GSEA

To perform GSEA with SomaEnrich, the following inputs are required:

1. A named vector of ranking statistics (`rank_vec`) with one measurement per gene
1. Biological networks or gene sets to interrogate for enrichment


### Selecting Gene Sets

The SomaEnrich package provides networks from both the Gene Ontology (GO) and
the Molecular Signatures Database (MSigDB). The default
network used in `somaPrGSEA()` is GO Biological Process (BP).
A full list of available networks and a short description of each collection
can be found via `?somaPrGSEA()`.

Please note that only _one_ biological network can be selected.
Mixing networks and/or subcollections can inflate and distort the null
distribution, because the gene sets differ wildly in size, redundancy, and
biological intent [@gsea-paper]. Additionally, selecting only one collection
helps keep processing time down, while also avoiding over-testing as a
consequence of performing too many statistical tests on large numbers of
gene sets.


### Putting it All Together

We will run pre-ranked GSEA with our `r rank_metric` vector, using GO BP for
enrichment.

```{r run-gsea, message=TRUE}
gsea_res <- somaPrGSEA(ranks = prepped_ranks,
                       resource = "bp")
```

A few messages will be produced when this function is run interactively:

* When `split_ids = TRUE` (the default), heterodimers are split into multiple values, one for each gene/protein in the complex, and this can create ties in the ranked input vector
* When `resolve_multimapping = TRUE` (the default), duplicate gene measurements will be filtered, selecting one value for retention (in cases of non-1:1 mapping) and other values will be dropped from the ranked input vector (Note: this may be outputted by `prepareRanks()` in a previous step)

This ensures that the names of the `rank_vec` are unique, and that only one
measurement per gene is used for GSEA.

The output contains two objects:

* A `data.frame` object containing the results of GSEA, with one row per pathway
* A vector of the ranks used as input into GSEA, which may differ from the input ranks (reference `prepareRanks()` for more details)

```{r view-gsea-results}
summary(gsea_res)

head(gsea_res$final_ranks)

gsea_res$results |>
  dplyr::select(-leadingEdge) |>
  head(n = 10)
```

Descriptions of what each column contains can be found via `?somaPrGSEA()`.
As expected, some of the top enriched pathways are sex-related. Column
descriptions for `pval`, `padj`, `ES`, and `NES` can be found in
`?somaPrGSEA()`.

Let's take a closer look at the `leadingEdge` column, which was removed from
the preview above:

```{r leading-edge-col}
filter(gsea_res$results, pathway_id == "GO:0022602") |>
    pull(leadingEdge)
```

The values in `leadingEdge` are the leading edge genes, i.e.
the core subset of features that account for the observed enrichment signal in
that gene set (see the 
[original GSEA software documentation](https://docs.gsea-msigdb.org/#GSEA/GSEA_User_Guide/#running-a-leading-edge-analysis)
for more details about leading edge analysis). So, in the example above, 
the leading edge genes were found to be concentrated toward the top of the
`rank_vec` ranked list for the pathway GO:0022602 ("ovulation cycle process").

Remember the sex-associated analytes in the `t_tests` data set that were
identified earlier in this vignette? Genes associated with these analytes, like
CGA and FSHB, were frequently identified as leading edge genes.

```{r leading-edge-count}
# Add column indicating if the leading edge subset contain CGA/FSHB
gsea_res$results |>
    mutate(cga_fshb_le = grepl("CGA|FSHB", leadingEdge)) |>
    select(pathway, leadingEdge, cga_fshb_le) |>
    head(n = 10)
```

These sex-associated genes drive a large portion of the enrichment identified
in each of these gene sets.

Next, let's look at the set size columns:

```{r set-size-col}
select(gsea_res$results, pathway, contains("size")) |>
    head()
```

These columns illustrate how many genes in each gene set were filtered from the
set, _prior_ to running GSEA. This is a standard pre-processing step for GSEA;
genes present in each gene set that are not found in the `ranks` vector are
removed prior to enrichment. This can have a small or large impact on the set
size, depending on the content of the set.


### AptNames Instead of Gene IDs

As mentioned in the previous section, GSEA can be run using SomaScan `AptNames`
rather than gene IDs, if desired. To do so, the vector provided to `ranks` must
be named with `AptNames`:

```{r create-apt-vec}
apt_vec <- t_tests$log2_fc
names(apt_vec) <- t_tests$AptName

head(apt_vec)
```

Note that `SeqIds` are *not* accepted as vector names - the R-compatible
`AptName` format (`seq.1234.56`) must be used. If starting from `SeqIds`, the
`SomaDataIO::seqid2apt()` function can be used to convert between the two
formats. See `?SomaDataIO::seqid2apt()` for more details.

Additionally, the column metadata from a `soma_adat` object will be required as
input. The information in the metadata will be used to map genes from GO and
MSigDB to SomaScan assay identifiers .

The column metadata can be extracted from a `soma_adat` using
`SomaDataIO::getAnalyteInfo()`:

```{r column-meta}
col_meta <- SomaDataIO::getAnalyteInfo(example_data_11k)

head(col_meta)
```

When running `somaPrGSEA()` with `AptNames`, is it important to set
`use_aptnames = TRUE` _and_ provide a data frame of column meta data to
`col_meta_df`. This will produce a progress bar that tracks the
conversion of GO BP terms from their original gene identifiers
to SomaScan `AptNames`, so gene set enrichment can be performed in aptamer
space (i.e. using `AptNames` in the ranking vector and `AptName`-containing
pathways).

Once all gene sets have been converted to `AptNames`, GSEA will be performed
using the ranked `AptNames`.

```{r run-apt-gsea, message=TRUE}
gsea_res_apt <- somaPrGSEA(ranks        = apt_vec,
                           col_meta_df  = col_meta, # Required
                           use_aptnames = TRUE)     # Required)

gsea_res_apt$results |>
  dplyr::select(resource_code:NES)
```

Because GSEA was performed in aptamer space, the `leadingEdge` column contains
leading edge **analytes**, rather than genes.

```{r apt-gsea-leading-edge}
gsea_res_apt$results |>
  dplyr::select(leadingEdge)
```


## Visualizing Results

`SomaEnrich` has a few built-in functions for visualizing ORA and GSEA results.
`plotES()` will create a plot of the enrichment score for any gene set,
using the results of `somaPrGSEA()`. The plot is very similar to that
produced by the Broad's original GSEA software, with the option to
annotate the leading edge genes on the enrichment score line.

Only one gene set can be plotted at a time. Let's use the same example gene
set as the previous section, "ovulation cycle":

```{r plot-gsea-res, fig.width=8, fig.height=6}
plotES("female gamete generation",
       gsea_results = gsea_res, 
       show_leading_edge = TRUE)
```

The dots annotated on the line at the top of the plot represent each of the
leading edge genes in the gene set.
When the NES value is positive, the leading edge genes are found at the left of
the plot. However, if the NES was negative, we'd see these genes on the right
side of the plot, _after_ the apex of the enrichment line.

```{r}
plotES("amino acid biosynthetic process",
       gsea_results = gsea_res,
       show_leading_edge = TRUE)
```


---------------

# Using Custom Gene Sets

If you want to use a biological network that is not already available in
SomaEnrich, you will need to:

1) Read in the network from a GMT file or other resource
1) Create a named list containing the new gene sets/pathways of interest

The output of the steps above can be provided to the `cust_paths` argument of
`somaPrGSEA()` or `somaORA()`. This section of the vignette will
describe how to do both steps above.


### GMT Files

A GMT file can easily be read in and provided to `somaPrGSEA()` using tools
from the `fgsea` package.

The `fgsea::gmtPathways()` function allows the user to read the contents of any
GMT file into R as a named list, with each list item corresponding to a
pathway from the GMT file. The example GMT file in SomaEnrich can be used to
illustrate this:

```{r ingest-gmt}
gmt_file <- system.file("extdata", "msigdb_c2_reactome.gmt",
                        package = "SomaEnrich")

gmt_list <- fgsea::gmtPathways(gmt_file)

head(gmt_list)
```

This list can be directly provided as-is to `somaPrGSEA()`:

```{r run-gsea-reactome}
gsea_cust <- somaPrGSEA(ranks = rank_vec,
                        cust_paths = gmt_list)

head(gsea_cust$results, n = 10)
```

These pathways can be visualized with the same functions as the built-in GO and
MSigDB resources. However, when using custom pathways, the `gmt_list` object
_must_ be provided to the `cust_path` argument:

```{r plot-cust-list, fig.width=7, fig.height=6}
plotES("REACTOME_G_ALPHA_S_SIGNALLING_EVENTS",
       gsea_res = gsea_cust,
       cust_path = gmt_list,
       show_leading_edge = TRUE)
```


## KEGG pathways

Currently, only GO and MSigDB networks are available in SomaEnrich. However,
other networks, like KEGG, that are not provided by SomaEnrich can still be
used as input to `SomaEnrich::somaPrGSEA()`. There are a number of tools
available to work with KEGG data in R; the example below will describe how
to use pathway analysis functions from the `limma` R package to retrieve KEGG
pathway information.

```{r load-kegg-content}
# Retrieve full list of available human KEGG pathways
kegg_content <- limma::getGeneKEGGLinks("hsa")

head(kegg_content)
```

This data frame must be converted to a list object for input into
`somaPrGSEA()`:

```{r make-kegg-list}
kegg_list <- df2list(kegg_content,
                     name_col  = "PathwayID",
                     value_col = "GeneID")

length(kegg_list)

head(kegg_list)
```

From this point, the `kegg_list` object can be provided to `somaPrGSEA()`
using the `cust_path` argument, as shown previously.

---------------

# Session Info

```{r session-info}
sessionInfo()
```


-------------

# References
